---
title: "tyramine_kymograph_analysis_final"
author: "Jacob Musser"
date: "`r doc_date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
        highlight: tango
        code_folding: show
    BiocStyle::pdf_document:
        toc: true
        highlight: tango
---

## Kymograph dissimilarity analysis (Fig 1K)

Quantifies canal activity and tissue movements in tyramine- vs DMSO-treated
*S. lacustris* from brightfield kymograph images. The primary metric is the
Pearson correlation distance (1 − *r*) between consecutive z-normalized frames,
measured at 1-minute intervals over 12 hours.

### Input files

- `microscopy/data/kymo_metadata.tsv` — Replicate metadata (file paths, treatment, replicate ID)
- `microscopy/data/*.csv` — Kymograph CSV files (rows = timepoints, cols = positions along the line)

### Output files

- `microscopy/outs/{date}_tyramine_kymograph_analysis/` — Correlation distance plot (Fig 1K) and time series CSV

# 1. Setup

## Rmarkdown knitr Settings
```{r setup, include = FALSE}
library(knitr)
options(digits = 3, width = 80)
opts_chunk$set(echo = TRUE,tidy = FALSE,include = TRUE,
               dev = 'png', comment = '  ',
               dpi = 300, cache.lazy = FALSE, warning = FALSE)
```

## Load Libraries
```{r load libraries, message = F, warning = F}
library(tidyverse)
library(here)

theme_set(theme_bw())
```

## Set User Parameters
```{r user parameters}

### User-adjustable parameters ###

metadata_file <- here("microscopy", "data", "kymo_metadata.tsv")

# Frame interval (units: minutes)
frame_interval <- 1

# Exclude pre-lid bright frames: original frames 1–11
start_frame <- 12L

### Output directory for plots ###

base_name  <- "tyramine_kymograph_analysis"
date_stamp <- format(Sys.Date(), "%Y%m%d")
output_dir <- here("microscopy", "outs", paste0(date_stamp, "_", base_name))

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
}

# Helper to save plots into the output directory
save_plot <- function(p, filename, width = 7, height = 5, dpi = 300) {
  ggsave(
    filename = file.path(output_dir, filename),
    plot     = p,
    width    = width,
    height   = height,
    dpi      = dpi
  )
}
```

## Functions
```{r functions}

### Load kymograph (rows = timepoints, cols = positions) ###

load_kymograph <- function(path) {
mat <- readr::read_csv(path, col_names = FALSE, show_col_types = FALSE)
as.matrix(mat)
}

### Z-normalize each frame across positions ###

z_normalize_frames <- function(I) {
mu  <- rowMeans(I, na.rm = TRUE)
sdv <- apply(I, 1, sd, na.rm = TRUE)
if (sum(sdv == 0)>0 | sum(is.na(sdv))>0) {
  print("SDV is 0 or NA")
}
sdv[sdv == 0 | is.na(sdv)] <- 1  # avoid division by zero
(I - mu) / sdv
}

### Correlation distance between consecutive z-normalized frames ###

compute_corr_distance <- function(I_z) {
n_time <- nrow(I_z)
if (n_time < 2) stop("Need at least two timepoints.")

d_corr <- rep(NA_real_, n_time - 1)
for (t in seq_len(n_time - 1)) {
  v1 <- I_z[t, ]
  v2 <- I_z[t + 1, ]
  if (all(is.na(v1)) || all(is.na(v2))) {
    d_corr[t] <- NA_real_
  } else {
    r <- suppressWarnings(stats::cor(v1, v2, use = "pairwise.complete.obs"))
    d_corr[t] <- 1 - r
  }
}

d_corr
}

# Matrix -> tidy tibble for kymographs

matrix_to_tidy <- function(M, time_vec = NULL) {
n_time <- nrow(M)
if (is.null(time_vec)) {
time_vec <- seq_len(n_time)
}

as_tibble(M) |>
mutate(
time_index = seq_len(n_time),
time       = time_vec
) |>
pivot_longer(
cols      = -c(time_index, time),
names_to  = "position_index",
values_to = "value"
) |>
mutate(
position_index = readr::parse_number(position_index)
)
}
```

## Per-replicate analysis function
```{r per replicate analysis function}

analyze_replicate <- function(file, treatment, replicate,
                              start_frame, frame_interval) {

  # ------------------------------------------------------------------
  # 1. Load raw and truncate at start_frame
  # ------------------------------------------------------------------
  I_full <- load_kymograph(file)
  I <- I_full[start_frame:nrow(I_full), , drop = FALSE]

  n_time <- nrow(I)
  time_vec <- (seq_len(n_time) - 1) * frame_interval

  # ------------------------------------------------------------------
  # 2. Z-normalize each frame across positions
  # ------------------------------------------------------------------
  I_z <- z_normalize_frames(I)

  # ------------------------------------------------------------------
  # 3. Correlation distance between consecutive frames
  # ------------------------------------------------------------------
  d_corr <- compute_corr_distance(I_z)
  time_diff <- time_vec[-1]

  # ------------------------------------------------------------------
  # 4. Return time series
  # ------------------------------------------------------------------
  ts <- tibble(
    file      = file,
    treatment = treatment,
    replicate = replicate,
    time      = time_diff,
    corr_dist = d_corr
  )

  ts
}

analyze_wrapper <- function(file, treatment, replicate) {
  analyze_replicate(
    file           = file,
    treatment      = treatment,
    replicate      = replicate,
    start_frame    = start_frame,
    frame_interval = frame_interval
  )
}
```

# 2. Read metadata and perform analysis on all files
```{r read metadata and perform analysis}
metadata <- readr::read_tsv(metadata_file, show_col_types = FALSE)
metadata$file <- here(metadata$file)

ts_all <- purrr::pmap_dfr(
  list(metadata$file, metadata$treatment, metadata$replicate),
  analyze_wrapper
)

cat("Replicates:", nrow(metadata), "\n")
cat("Treatment groups:", paste(names(table(metadata$treatment)),
    table(metadata$treatment), sep = "=", collapse = ", "), "\n")

write.csv(ts_all, file.path(output_dir, "ts_all.csv"), row.names = FALSE)
```

# 3. Correlation distance plot (Fig 1K)

```{r theme}
custom_theme <- theme(
  text = element_text(family = "sans", size = 16),
  plot.title = element_text(size = 18, face = "bold"),
  axis.title = element_text(size = 16),
  axis.text = element_text(size = 12),
  legend.title = element_blank(),
  legend.text = element_text(size = 10),
  legend.key.size = unit(0.8, "lines"),
  legend.background = element_rect(fill = "white", color = "gray80", linewidth = 0.3),
  panel.grid = element_blank()
)
```

```{r correlation distance ribbon plot}
# Capitalize treatment labels for legend
ts_all$treatment[ts_all$treatment == "tyramine"] <- "Tyramine"

p_corr_all <- ggplot(ts_all,
                     aes(x = time, y = corr_dist,
                         color = treatment, fill = treatment,
                         group = treatment)) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.3, color = NA) +
  stat_summary(fun = mean, geom = "line", size = 0.75) +
  scale_color_manual(values = c("red", "blue"), labels = c("DMSO", "Tyramine")) +
  scale_fill_manual(values = c("red", "blue"), labels = c("DMSO", "Tyramine")) +
  labs(
    x = "Time (min)",
    y = bquote(1 - rho ~ "(" ~ I[ Z ~ t ] * " ," ~ I[ Z ~ t+1 ] ~ ")" )
    )+
  custom_theme +
  theme(legend.position = c(0.9, 0.9)) +
  scale_x_continuous(limits = c(-1, 721), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 0.15), expand = c(0, 0))

p_corr_all
save_plot(p_corr_all, "correlation_distance_fig1K.png")
```

```{r session info}
sessionInfo()
```
